Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-03 paper/paper.tex
Update entrega

\setstretch{1.5}
This document was last modified at 2025-11-03 01:00:05 and it was last
rendered at 2025-11-03 01:00:05.

@@

Social mobility effects---understood as impacts on an outcome arising
from movements between an origin state and a destination state (e.g.,
social class)---have long been a focus of sociological research
(\citeproc{ref-eyles_social_2022}{Eyles et al., 2022};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
Yet, as Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024, p.
467}) emphasize, most hypotheses are, at their core, individual-level
counterfactual comparisons between the observed outcome under a mobility
trajectory and the outcome the same person would have shown if they had
remained in their origin class (or moved to an alternative destination).
Standard specifications (e.g., SAM, DRM, and mobility-contrast models)
struggle to retrieve these inherently counterfactual quantities due to
identification constraints and their reliance on between-group contrasts
constructed from additive terms and interactions, thereby yielding
primarily associational evidence
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). I therefore adopt
Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024}) causal
formulation, treating destination class as the treatment, conditioning
on the class of origin, and estimating heterogeneous destination effects
with observational data under explicit identification assumptions. To
align design and target, I follow the MIDA template
(\citeproc{ref-blair_research_2023}{Blair et al., 2023}): set out the
causal model and assumptions (M), define the inquiry and estimand (I),
describe the data and variables (D), and detail the answer strategy (A)
used to identify the causal effect of intergenerational social mobility
on preferences for the commodification of pensions.

@@

\textbf{Positivity}: For all \((j,c)\) in the support of \((O,C)\) and
for each relevant destination \((k)\), the probability of receiving
\(D=k\) is strictly between 0 and 1; substantively, every \((O,C)\)-type
has non-zero chance of each comparison destination.

@@

In fact, establishing the causal relationship between social mobility
and subjective outcomes, such as preferences, poses several challenges
for causal inference. Below, I use directed acyclic graphs (DAGs) to
illustrate some of these challenges and evaluate possible strategies for
identifying the effects of social mobility. Figure~\ref{fig-dag} depicts
the causal model guiding identification. In this model, a person's
preference (\(Y\)) is directly caused by her class destination (\(D\))
and can also be affected by different attributes determined in childhood
or earlier (\(C\)), such as family resources, household composition, and
ascriptive status (pre-treatment confounders). Class origin (\(O\)) is
determined by those origin-attributes (\(C\)) and, in turn, affects
destination (\(D\)). In addition, denotes unobserved factors (\(U\)) can
also influence origin-attributes (\(C\)). There is no direct arrow from
\(O\) to \(Y\), reflecting the assumption that the effect of origin on
the outcome operates entirely through background factors \((C)\) and
destination \((D)\). Accordingly, conditioning on \((O,C)\) could blocks
the relevant back-door paths from \(D\) to \(Y\), specifically
\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
adjustment set.

@@

\caption{\label{fig-dag}Causal graph of the effect of intergenerational
social mobility on preferences for commodification. Y = preferences for
pension commodification, my outcome of interest, D = an individual's
class of destination, O = an individual's class of origin, C = different
attributes determined in childhood or earlier that affects an
individual's class of origin and destination. Finally, U = unobserved
factors influencing origin attributes.}

@@

This structure corresponds to a nonparametric structural causal model in
which \((D)\) is the treatment variable, \((Y)\) the outcome, and
\((O, C)\) the minimal sufficient adjustment set. Conditioning on these
variables satisfies the back-door criterion for identifying the causal
effect of mobility on preferences
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024}).

Why are the assumptions plausible here? Conditional independence is
targeted by controlling for a plausible sufficient adjustment set
\((O,C)\); entropy balancing/IPW enforces covariate balance within each
origin stratum. Positivity is assessed via propensity scores and
trimming of extreme weights, restricting to common support. SUTVA holds
by treating the destination class as a single exposure measured before
\((Y)\) and assuming no interference. Together, the DAG and assumptions
define conditions for identifying the causal effect of intergenerational
mobility on preferences for commodification.

@@

To test heterogeneity in mobility effects, I examine two moderators:
meritocratic perceptions and time. Meritocracy is captured with two
items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
effort (``In Chile, people are rewarded for their efforts'') and one to
talent (``In Chile, people are rewarded for their intelligence and
skills''), each answered on a five-point Likert scale (1 = ``strongly
disagree'' to 5 = ``strongly agree''). I average the two items into a
single index and dichotomize it into low meritocracy (≤ 3) and high
meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
and entered as a categorical (dummy) variable. Both variables are used
as moderators by interacting them with the mobility treatment to assess
conditional average treatment effects---i.e., whether the effect of
mobility varies across levels of meritocratic beliefs or across survey
waves.

@@

where \(Y_{it}\) indicates whether individual \(i\) in wave \(t\)
supports more market-based pension access; \(\alpha\) is the baseline
probability for the reference categories; \(\beta T_i\) captures the
intergenerational mobility contrast: \(T_i=1\) when the observed
destination is \(D_i=k\) (mobile to \(k\)) and \(T_i=0\) when \(D_i=j\)
(immobile), so that \(\beta\) is the ATT within origin
\(O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})\), i.e., the
percentage-point change in the probability of preferring pension
commodification from reaching \(k\) rather than remaining in \(j\). The
term \(X_i'\gamma\) includes the pre-treatment covariates \((C)\) used
to construct the entropy-balancing weights; \(\lambda_t\) are wave fixed
effects; and \(\varepsilon_{it}\) is idiosyncratic error. The analytic
sample is restricted to individuals sharing the same origin \((O=j)\).
Estimation uses pooled WLS with stabilized IPW--ATT from entropy
balancing and CR2 standard errors clustered by individual to address
heteroskedasticity and within-person dependence.

@@

Figure~\ref{fig-alluvial} shows the annual frequencies of preferences
for pension commodification in 2016, 2018, and 2023. Each year presents
stacked percentages frequencies for the level of agreement and
disagreement. Overall, a large majority rejects market-based access to
old-age benefits, but this opposition has eased over time: 83.5\% in
2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018 change is
modest, whereas 2023 registers a 9.4 point drop in disagreement relative
to 2018, mirrored by rising agreement: 16.5\% (2016), 18.8\% (2018),
28.2\% (2023). Substantively, while most respondents continue to oppose
the idea that higher-income individuals should obtain better pensions
via the market, a non-trivial and growing portion endorses
commodification, with the sharpest expansion concentrated in the latest
wave (+9.4 points from 2018 to 2023).

@@

Regarding the relationship between preferences for the commodification
of pensions and intergenerational occupational mobility,
Figure~\ref{fig-mean} shows the percentage of agreement with the market
provision of pension benefits according to mobility trajectories and
survey waves. Averaging across waves, agreement is highest among the
High-High immobile (27.8\%), followed by the Middle-High upwardly mobile
(24.6\%) and the High-Middle downwardly mobile (24.0\%); the lowest
levels are found among Low-High (17.3\%), Low-Low (17.4\%), and
descending High-Low (17.6\%) groups. The specific profiles of each wave
accentuate these gradients, especially in 2023: agreement reaches 31.9\%
for high-high, 29.4\% for upward middle-high, and peaks at 36.8\% for
downward high-middle (the highest of all trajectories for that year).
Descriptively, immobility at the top is associated with greater support
for commodification; among those who move, support is greatest for the
upward Middle-High trajectory and the downward High-Middle trajectory,
patterns that intensify in 2023. These shifts anticipate the
heterogeneity documented below.

@@

Figure~\ref{fig-reg} reports the doubly robust estimates of the effect
of intergenerational mobility on preferences for pension
commodification, obtained from weighted models combining stabilized IPW
with covariate adjustment and wave fixed effects. Results reveal a clear
directional asymmetry between upward and downward mobility trajectories.
Among upwardly mobile respondents, only the Middle→High trajectory
exhibits a significant positive effect on support for pension
commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
Interpreted as an origin-specific ATT, this estimate compares---for
individuals who actually moved from the middle to the high class---their
observed endorsement with the counterfactual endorsement they would have
expressed had they remained immobile in the middle class. Holding all
other predictors constant, the point estimate implies a 9 percentage
point higher probability of endorsing market-based access to pension
benefits for Middle→High movers relative to their own immobility
counterfactual, statistically significant at the 95\% confidence level.
In contrast, two downward trajectories: High→Middle (\(\beta\) = −0.07,
CI {[}−0.15, −0.00{]}) and High→Low (\(\beta\) = −0.10, CI {[}−0.18,
−0.02{]}), show significant negative effects, suggesting that
individuals who experience downward mobility from higher-class origins
express weaker preferences for market provision. The remaining pathways
(Low→Middle, Low→High, Middle→Low) display no significant differences
relative to their non-mobile counterparts. These patterns indicate that
upward movement within the upper segment (Middle→High) reinforces
pro-commodification attitudes, whereas downward movement from privileged
origins reduces them.

@@

Figure~\ref{fig-interact} shows that neither the interaction terms nor
the mobility coefficients reach conventional significance across
trajectories, indicating no detectable moderation of the causal mobility
effect by meritocratic perceptions at the 95\% level. Directionally, the
interaction is positive for Middle→High (\(\beta\) = 0.15, 95\% CI
{[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05, {[}−0.14, 0.24{]}),
suggesting, if anything, somewhat larger pro-commodification effects
among high-merit respondents when upwardly mobile, whereas it is
negative for High→Low (\(\beta\) = −0.14, {[}−0.38, 0.10{]}) and
High→Middle (\(\beta\) = −0.03, {[}−0.23, 0.17{]}), showing weaker
pro-commodification effects under downward mobility at high merit (left
panel). Focusing on the treatment effect at low meritocracy (the
reference category), the mobility coefficient captures the
origin-specific ATT comparing movers to their immobility counterfactual
among low-merit respondents (right panel). These effects are small and
imprecise across trajectories: Low→Middle (\(\beta\) = −0.02, {[}−0.09,
0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]}) are near
zero. Middle→High is positive but not significant (\(\beta\) = 0.07,
{[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00, {[}−0.07,
0.06{]}), and downward from high shows negative, borderline estimates:
High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and High→Low
(\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not find
evidence that meritocratic perceptions moderate the causal effect of
mobility, although the signs align with the idea that upward mobility at
high merit tilts attitudes toward commodification, and downward mobility
at high merit tilts them away.

@@

Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
effect across the different waves of the survey. The treatment effect in
2016 (reference category) closely replicates the reference ATT patterns:
positive for Middle→High (\(\beta\) = 0.13, 95\% CI {[}0.05, 0.22{]})
and negative for High→Middle (\(\beta\) = −0.19, {[}−0.33, −0.05{]}) and
High→Low (\(\beta\) = −0.19, {[}−0.32, −0.05{]}). This indicates that,
holding all other predictors constant, individuals who moved from Middle
to high status in 2016 were, on average, 13 percentage points more
likely to support access to the pension market compared to their
counterfactual of immobility, while those who moved down from higher
classes were approximately 19 percentage points less likely. Beyond
these reference effects, there is a notable temporal change in the
High→Middle trajectory: the interaction between mobility and the 2023
wave is positive and statistically significant (\(\beta\) = 0.19,
{[}0.01, 0.38{]}). This suggests that, although people with downward
mobility from high backgrounds were less supportive of commodification
in 2016, by 2023 the same trajectory shows a change in sign, implying
that the negative causal effect of downward mobility attenuated and
eventually turned positive. No comparable changes are observed in other
trajectories. In essence, this pattern reveals time-conditional
heterogeneity in the mobility effect: among those who experienced
downward mobility, preferences for market-based access to old-age
pensions appear to have realigned in the most recent wave, reflecting a
notable shift toward greater support for commodification within this
trajectory by 2023.

@@

The first sensitivity check corresponds to Ding \& VanderWeele
(\citeproc{ref-ding_sensitivity_2016}{2016}) approach to assess
unmeasured confounding (E values). This indicates the minimum strength
of association that an unobserved confounder would need to have with
both the treatment and the outcome---on the relative risk scale---to
explain an effect, the Middle-High ATT would require an \(RR ≈ 1.74\) to
offset the point estimate (\(E_{point}\) = 1.736) and an \(RR ≈ 1.26\)
to move its CI to include the null value (\(E_{CI}\) = 1.264). High→Low
shows comparable robustness (\(E_{point}\) = 1.759; \(E_{CI}\) = 1.239),
while High→Middle is comparatively fragile near zero (\(E_{point}\) =
1.591; \(E_{CI}\) = 1.037). For non-significant trajectories
(Low→Middle, Low→High, Middle→Low), \(E_{CI}\) = 1, as expected. In
conclusion, the directional pattern---stronger in favor of
commodification in Middle→High and weaker in the descent from
High---seems reasonably robust to moderate unmeasured confounding in
Middle→High and High→Low, but borderline in High→Middle. These results
justify treating signs and relative magnitudes as informative, while
recognizing limited protection against omitted variables for High→Middle
(see \hyperref[anexo]{Supplementary Material} for complete tables).

The second sensitivity check applies Holm's step-down correction---a
multiple-testing procedure that controls the familywise error rate
(FWER) while retaining more power than Bonferroni---across the six
mobility trajectories. After FWER control, none of the effects remain
significant at \(\alpha\) = .05: Middle→High (\(p_{raw}\) = .0116 →
\(p_{Holm}\) = .0698), High→Low (.0167 → .0834), and High→Middle (.0481
→ .1924). Accordingly, claims of statistical significance should be
tempered under strict multiplicity control. Substantively, however, this
adjustment affects the strength of inference, not its direction: the
pattern remains consistent with the theorized asymmetry (positive for
Middle→High; negative for downward from high). I therefore treat signs
and effect sizes as the primary evidence and regard p-values as
exploratory once multiplicity is accounted for (see
\hyperref[anexo]{Supplementary Material} for complete tables).

>>>>

\setstretch{1.5}
This document was last modified at 2025-11-03 17:04:08 and it was last
rendered at 2025-11-03 17:04:08.

@@

Social mobility effects---understood as impacts on an outcome arising
from movements between an origin state and a destination state (e.g.,
social class)---have long been a focus of sociological research
(\citeproc{ref-eyles_social_2022}{Eyles et al., 2022};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}).
Yet, as Breen and Ermisch (\citeproc{ref-breen_effects_2024}{2024, p.
467}) emphasize, most of the mobility hypotheses are, at their core,
individual-level counterfactual comparisons between the observed outcome
under a mobility trajectory and the outcome the same person would have
shown if they had remained in their origin class (or moved to an
alternative destination). Standard specifications (e.g., SAM, DRM, and
mobility-contrast models) struggle to retrieve these inherently
counterfactual quantities due to identification constraints and their
reliance on between-group contrasts constructed from additive terms and
interactions, thereby yielding primarily associational evidence
(\citeproc{ref-breen_effects_2024}{Breen \& Ermisch, 2024};
\citeproc{ref-song_there_2025}{Song \& Zhou, 2025}). I therefore adopt
Breen and Ermisch's (\citeproc{ref-breen_effects_2024}{2024}) causal
formulation, treating destination class as the treatment, conditioning
on the class of origin, and estimating heterogeneous destination effects
with observational data under explicit identification assumptions. To
align design and target, I follow the MIDA template
(\citeproc{ref-blair_research_2023}{Blair et al., 2023}): set out the
causal model and assumptions (M), define the inquiry and estimand (I),
describe the data and variables (D), and detail the answer strategy (A)
used to identify the causal effect of intergenerational social mobility
on preferences for the commodification of pensions.

@@

\textbf{Positivity}: For all \((j,c)\) in the support of \((O,C)\) and
for each relevant destination \((k)\), the probability of receiving
\(D=k\) is strictly between 0 and 1; substantively, each type \((O,C)\)
has a nonzero probability of entering each comparison destination.

@@

In fact, establishing the causal relationship between objective social
mobility and subjective outcomes, such as preferences, poses several
challenges for causal inference. Below, I use directed acyclic graphs
(DAGs) to illustrate some of these challenges and evaluate possible
strategies for identifying the effects of social mobility.
Figure~\ref{fig-dag} depicts the causal model guiding identification. In
this model, a person's preference (\(Y\)) is directly caused by her
class destination (\(D\)) and can also be affected by different
attributes determined in childhood or earlier (\(C\)), such as family
resources, household composition, and ascriptive status (pre-treatment
confounders). Class origin (\(O\)) is determined by those
origin-attributes (\(C\)) and, in turn, affects destination (\(D\)). In
addition, unobserved factors (\(U\)) can also influence
origin-attributes (\(C\)). There is no direct arrow from \(O\) to \(Y\),
reflecting the assumption that the effect of origin on the outcome
operates entirely through background factors \((C)\) and destination
\((D)\). Accordingly, conditioning on \((O,C)\) could blocks the
relevant back-door paths from \(D\) to \(Y\), specifically
\(D\leftarrow O\leftarrow C\rightarrow Y\) and, if present,
\(D \leftarrow O \rightarrow Y\), so \((O,C)\) is a minimal sufficient
adjustment set.

@@

\caption{\label{fig-dag}Causal graph of the effect of intergenerational
social mobility on preferences for commodification. Y = preferences for
pension commodification, D = an individual's class of destination, O =
an individual's class of origin, C = different attributes determined in
childhood or earlier that affects an individual's class of origin and
destination. Finally, U = unobserved factors influencing origin
attributes.}

@@

This structure corresponds to a nonparametric structural causal model in
which \(D\) is the treatment variable, \(Y\) the outcome, and \((O, C)\)
the minimal sufficient adjustment set. Conditioning on these variables
satisfies the back-door criterion for identifying the causal effect of
mobility on preferences (\citeproc{ref-breen_effects_2024}{Breen \&
Ermisch, 2024}).

Why are the assumptions plausible here? Conditional independence is
targeted by controlling for a plausible sufficient adjustment set
\((O,C)\); inverse probability weights (IPW) obtained via entropy
balancing enforces covariate balance within each origin stratum.
Positivity is assessed via propensity scores and trimming of extreme
weights, restricting to common support. SUTVA holds by treating the
destination class as a single exposure measured before \((Y)\) and
assuming no interference. Together, the DAG and assumptions define
conditions for identifying the causal effect of intergenerational social
mobility on preferences for commodification.

@@

To test heterogeneity in mobility effects, I examine two moderators:
meritocratic perceptions and time. Meritocracy is captured with two
items (\citeproc{ref-young_rise_1958}{Young, 1958}), one referring to
effort (``In Chile, people are rewarded for their efforts'') and one to
talent (``In Chile, people are rewarded for their intelligence and
skills''), each answered on a five-point Likert scale (1 = ``strongly
disagree'' to 5 = ``strongly agree''). I average the two items into a
single index and dichotomize it into low meritocracy (≤ 3) and high
meritocracy (≥ 4). Time is measured by survey wave (2016, 2018, 2023)
and entered as a categorical (dummy) variable. Both variables are used
as moderators by interacting them with the mobility treatment to assess
conditional average treatment effects---i.e., whether the effect of
mobility varies across levels of meritocratic perceptions or across
survey waves.

@@

where \(Y_{it}\) indicates whether individual \(i\) in wave \(t\)
supports more market-based pension access; \(\alpha\) is the baseline
probability for the reference categories; \(\beta T_i\) captures the
intergenerational mobility contrast: \(T_i=1\) when the observed
destination is \(D_i=k\) (mobile to \(k\)) and \(T_i=0\) when \(D_i=j\)
(immobile), so that \(\beta\) is the ATT within origin
\(O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})\), i.e., the
percentage-point change in the probability of preferring pension
commodification from reaching \(k\) rather than remaining in \(j\). The
term \(X_i'\gamma\) includes the pre-treatment covariates \((C)\) used
to construct the entropy-balancing weights; \(\lambda_t\) are wave fixed
effects; and \(\varepsilon_{it}\) is idiosyncratic error. The analytic
sample is restricted to individuals sharing the same origin \((O=j)\).
Therefore, estimation uses pooled WLS with stabilized IPW--ATT from
entropy balancing and CR2 standard errors clustered by individual to
address heteroskedasticity and within-person dependence.

@@

Figure~\ref{fig-alluvial} shows the annual frequencies of preferences
for pension commodification in 2016, 2018, and 2023. Each year presents
stacked percentages frequencies for the level of agreement and
disagreement. Overall, a large majority rejects market-based access to
old-age pension benefits, but this opposition has eased over time:
83.5\% in 2016, 81.2\% in 2018, and 71.8\% in 2023. The 2016--2018
change is modest, whereas 2023 registers a 9.4 point drop in
disagreement relative to 2018, mirrored by rising agreement: 16.5\%
(2016), 18.8\% (2018), 28.2\% (2023). Substantively, while most
respondents continue to oppose the idea that higher-income individuals
should obtain better pensions via the market, a non-trivial and growing
portion endorses this statement, with the sharpest expansion
concentrated in the latest wave (+9.4 points from 2018 to 2023).

@@

Regarding the relationship between preferences for the commodification
of pensions and intergenerational occupational mobility,
Figure~\ref{fig-mean} shows the percentage of agreement with
market-based access to pension benefits according to mobility
trajectories and survey waves. Averaging across waves, agreement is
highest among the High-High immobile (27.8\%), followed by the
Middle-High upwardly mobile (24.6\%) and the High-Middle downwardly
mobile (24.0\%). The lowest levels are found among ascending Low-High
(17.3\%), and immobile Low-Low (17.4\%) groups. The specific profiles of
each wave accentuate these gradients in most trajectories, especially in
2023: agreement reaches 31.9\% for High-High, 29.4\% for upward
Middle-High, and peaks at 36.8\% for downward High-Middle (the highest
of all trajectories for that year). Descriptively, immobility at the top
is associated with greater support for commodification; among those who
move, support is greatest for the upward Middle-High trajectory and the
downward High-Middle trajectory, patterns that intensify in 2023. These
shifts anticipate the heterogeneity documented below.

@@

Figure~\ref{fig-reg} reports the double-robust estimates of the effect
of intergenerational occupational mobility on preferences for pension
commodification, obtained from weighted models that combine stabilized
IPW with covariate adjustment and fixed wave effects. Results reveal a
clear directional asymmetry between upward and downward mobility
trajectories. Among upwardly mobile respondents, only the Middle→High
trajectory exhibits a significant positive effect on support for pension
commodification (\(\beta\) = 0.09, 95\% CI {[}0.02, 0.16{]}).
Interpreted as an origin-specific ATT, this estimate compares---for
individuals who actually moved from the middle to the high socioeconomic
status---their observed endorsement with the counterfactual endorsement
they would have expressed had they remained immobile in the middle
socioeconomic status. Holding all other predictors constant, the point
estimate implies a 9 percentage point higher probability of endorsing
market-based access to pension benefits for Middle→High movers relative
to their own immobility counterfactual, statistically significant at the
95\% confidence level. In contrast, two downward trajectories:
High→Middle (\(\beta\) = −0.07, CI {[}−0.15, −0.00{]}) and High→Low
(\(\beta\) = −0.10, CI {[}−0.18, −0.02{]}), show significant negative
effects, suggesting that individuals who experience downward mobility
from higher-status origins express weaker preferences for market-based
pension benefits. The remaining pathways (Low→Middle, Low→High,
Middle→Low) display no significant differences relative to their
non-mobile counterparts. These patterns indicate that upward movement
within the upper segment (Middle→High) reinforces pro-commodification
attitudes, whereas downward movement from privileged origins reduces
them.

@@

Figure~\ref{fig-interact} shows that neither the interaction terms nor
the mobility coefficients reach conventional significance across
trajectories, indicating no detectable moderation of the causal mobility
effect by meritocratic perceptions at the 95\% level. Directionally, in
the left panel, the interaction is positive for Middle→High (\(\beta\) =
0.15, 95\% CI {[}−0.03, 0.33{]}) and Low→High (\(\beta\) = 0.05,
{[}−0.14, 0.24{]}), suggesting, if anything, somewhat larger
pro-commodification effects among high-merit respondents when upwardly
mobile, whereas it is negative for High→Low (\(\beta\) = −0.14,
{[}−0.38, 0.10{]}) and High→Middle (\(\beta\) = −0.03, {[}−0.23,
0.17{]}), showing weaker pro-commodification effects under downward
mobility at high merit. Focusing on the treatment effect at low
meritocracy (the reference category) at the right panel, the mobility
coefficient captures the origin-specific ATT comparing movers to their
immobility counterfactual among low-merit respondents. These effects are
small and imprecise across trajectories: Low→Middle (\(\beta\) = −0.02,
{[}−0.09, 0.04{]}) and Low→High (\(\beta\) = −0.03, {[}−0.10, 0.05{]})
are near zero. Middle→High is positive but not significant (\(\beta\) =
0.07, {[}−0.01, 0.14{]}), Middle→Low is null (\(\beta\) = -0.00,
{[}−0.07, 0.06{]}), and downward from high shows negative, borderline
estimates: High→Middle (\(\beta\) = −0.07, {[}−0.15, 0.00{]}) and
High→Low (\(\beta\) = −0.08, {[}−0.15, 0.00{]}). Substantively, I do not
find evidence that meritocratic perceptions moderate the causal effect
of mobility, although the signs align with the idea that upward mobility
at high merit tilts attitudes toward commodification, and downward
mobility at high merit tilts them away.

@@

Table~\ref{tbl-inttime} shows the causal heterogeneity of the mobility
effect across the different waves of the survey. The treatment effect in
2016 (reference category) closely replicates the reference ATT patterns:
positive for Middle→High (\(\beta\) = 0.13, 95\% CI {[}0.05, 0.22{]})
and negative for High→Middle (\(\beta\) = −0.19, {[}−0.33, −0.05{]}) and
High→Low (\(\beta\) = −0.19, {[}−0.32, −0.05{]}). This indicates that,
holding all other predictors constant, individuals who moved from middle
to high socioeconomic status in 2016 were, on average, 13 percentage
points more likely to support market-based access to the pension
benefits compared to their counterfactual of immobility, while those who
moved down from higher status were approximately 19 percentage points
less likely. Beyond these reference effects, there is a notable temporal
change in the High→Middle trajectory: the interaction between mobility
and the 2023 wave is positive and statistically significant (\(\beta\) =
0.19, {[}0.01, 0.38{]}). This suggests that, although people with
downward mobility from high backgrounds were less supportive of
commodification in 2016, by 2023 the same trajectory shows a change in
sign, implying that the negative causal effect of downward mobility
attenuated and eventually turned positive. No comparable changes are
observed in other trajectories. In essence, this pattern reveals
time-conditional heterogeneity in the mobility effect: among those who
experienced downward mobility, preferences for market-based access to
old-age pensions appear to have realigned in the most recent wave,
reflecting a notable shift toward greater support for commodification
within this trajectory by 2023.

@@

The first sensitivity check corresponds to the Ding \& VanderWeele
(\citeproc{ref-ding_sensitivity_2016}{2016}) approach to assess
unmeasured confounding (E values). This values indicates the minimum
strength of association that an unobserved confounder would need to have
with both the treatment and the outcome---on the relative risk
scale---to explain an effect. The Middle-High ATT would require an
\(RR ≈ 1.74\) to offset the point estimate (\(E_{point}\) = 1.736) and
an \(RR ≈ 1.26\) to move its CI to include the null value (\(E_{CI}\) =
1.264). High→Low shows comparable robustness (\(E_{point}\) = 1.759;
\(E_{CI}\) = 1.239), while High→Middle is comparatively fragile near
zero (\(E_{point}\) = 1.591; \(E_{CI}\) = 1.037). For non-significant
trajectories (Low→Middle, Low→High, Middle→Low), \(E_{CI}\) = 1, as
expected. In conclusion, the directional pattern---stronger in favor of
commodification in Middle→High and weaker in the descent from
High---seems reasonably robust to moderate unmeasured confounding in
Middle→High and High→Low, but borderline in High→Middle. These results
justify treating signs and relative magnitudes as informative, while
recognizing limited protection against omitted variables for High→Middle
(see \hyperref[anexo]{Supplementary Material} for complete tables).

The second sensitivity check applies Holm's step-down correction, which
is a multiple-testing procedure that controls the familywise error rate
(FWER) while retaining more power than Bonferroni, across the six
mobility trajectories. After FWER control, none of the effects remain
significant at \(\alpha\) = .05: Middle→High (\(p_{raw}\) = .0116 →
\(p_{Holm}\) = .0698), High→Low (.0167 → .0834), and High→Middle (.0481
→ .1924). Accordingly, claims of statistical significance should be
tempered under strict multiplicity control. Substantively, however, this
adjustment affects the strength of inference, not its direction: the
pattern remains consistent with the theorized asymmetry (positive for
Middle→High; negative for downward from high). I therefore treat signs
and effect sizes as the primary evidence and regard p-values as
exploratory once multiplicity is accounted for (see
\hyperref[anexo]{Supplementary Material} for complete tables).
